---
title: "Setting up Clinical Studies with gsm.datasim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Setting up Clinical Studies with gsm.datasim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup, eval = TRUE}
library(gsm.datasim)
```

# Introduction

The `gsm.datasim` package provides a comprehensive framework for generating synthetic clinical trial data. This vignette demonstrates how to set up both single snapshot and longitudinal studies, covering the key functions and configuration options available.

## Overview of Study Types

The package supports two main types of studies:

1. **Single Snapshot Studies**: Generate data for a single time point
2. **Longitudinal Studies**: Generate data across multiple time points to simulate study progression

# Single Snapshot Studies

A single snapshot study generates clinical data for one specific time point. This is useful for testing analytics workflows or when you need a static dataset.

## Basic Single Snapshot Setup

The simplest way to create a single snapshot study is using the study configuration system:

```{r single-basic}
# Create a basic study configuration
config <- create_study_config(
  study_id = "STUDY001",
  participant_count = 100,
  site_count = 5
)

# Set to a single snapshot (the default is 5)
config <- set_temporal_config(config, snapshot_count = 1)

# Add clinical domains (datasets)
config <- add_dataset_config(config, "Raw_AE", enabled = TRUE) # Adverse Events
config <- add_dataset_config(config, "Raw_LB", enabled = TRUE) # Laboratory Data
config <- add_dataset_config(config, "Raw_VISIT", enabled = TRUE) # Visit Data

# Generate the study data
raw_data <- generate_study_data(config, verbose = TRUE)

# Examine the generated data structure
names(raw_data)
```

## Configuring Temporal Settings

Even for single snapshots, you can control the temporal aspects:

```{r single-temporal}
# Create configuration with specific temporal settings
config <- create_study_config(
  study_id = "STUDY002",
  participant_count = 200,
  site_count = 10
)

# Set temporal configuration
config <- set_temporal_config(
  config,
  start_date = "2023-06-01",
  snapshot_count = 1, # Single snapshot
  snapshot_width = "months"
)

# Add domains with custom configurations
config <- add_dataset_config(config, "Raw_AE", enabled = TRUE)
config <- add_dataset_config(config, "Raw_LB",
  enabled = TRUE,
  growth_pattern = "exponential"
)

# Generate data
raw_data <- generate_study_data(config, verbose = TRUE)
```

## Adding Analytics Pipeline

You can run analytics on your single snapshot study. Results are returned from the `workr`-powered pipeline — a named list where each snapshot contains `$results` (named by workflow, e.g. `Analysis_kri0001`), `$mapped`, `$lWorkflow`, and `$summary`:

```{r single-analytics}
# Create study with analytics
config <- create_study_config(
  study_id = "STUDY003",
  participant_count = 150,
  site_count = 8,
  analytics_package = "gsm.kri"
)

# Add clinical domains
config <- add_dataset_config(config, "Raw_AE", enabled = TRUE)
config <- add_dataset_config(config, "Raw_LB", enabled = TRUE)

# Generate data
raw_data <- generate_study_data(config, verbose = TRUE)

# Run analytics pipeline — returns workr results per snapshot
analytics_results <- generate_analytics_layers(raw_data, config, verbose = TRUE)

# Each snapshot contains $results, $mapped, $lWorkflow, $summary
names(analytics_results[[1]])

# Access a specific metric result across all its data frames
kri_result <- analytics_results[[1]]$results$Analysis_kri0001
names(kri_result) # e.g. Analysis_Summary, Analysis_Flagged, Analysis_Analyzed
```

## Adding the Reporting Pipeline

The reporting pipeline runs reporting workflows on top of the analytics results. Pass the analytics output directly to `generate_reporting_layers()`:

```{r single-reporting}
# Run reporting pipeline on analytics results
reporting_results <- generate_reporting_layers(
  analytics_results = analytics_results,
  config = config,
  verbose = TRUE
)

# Results are keyed by snapshot name, matching analytics_results
names(reporting_results)
```

# Longitudinal Studies  

Longitudinal studies simulate data evolution over time, which is essential for understanding how clinical metrics change throughout a study.

## Basic Longitudinal Study

The `create_longitudinal_study()` function provides the easiest way to set up a longitudinal study. By default only raw data is generated; set `run_analytics = TRUE` to run the analytics pipeline and `run_reporting = TRUE` to also run the reporting pipeline:

```{r longitudinal-basic}
# Create a longitudinal study with default settings
study <- create_longitudinal_study(
  study_id = "LONG-001",
  participants = 100,
  sites = 5,
  snapshots = 6, # 6 time points
  interval = "1 month", # Monthly snapshots
  domains = c("AE", "LB", "VISIT"), # Clinical domains to include
  run_analytics = TRUE, # Run analytics pipeline
  run_reporting = TRUE, # Also run reporting pipeline
  verbose = TRUE
)

# Examine the study structure
names(study) # includes $raw_data, $analytics, $reporting

# View snapshot dates
names(study$raw_data)

# Analytics results: workr output per snapshot
if (!is.null(study$analytics)) {
  names(study$analytics[[1]]$results) # e.g. "Analysis_kri0001", ...
}

# Reporting results per snapshot
if (!is.null(study$reporting)) {
  names(study$reporting)
}
```

## Advanced Longitudinal Configuration

For more control over the longitudinal study setup:

```{r longitudinal-advanced}
# Create a complex longitudinal study
study <- create_longitudinal_study(
  study_id = "ADVANCED-001",
  participants = 500,
  sites = 25,
  snapshots = 12, # 1 year of monthly data
  interval = "1 month",
  domains = c("AE", "LB", "VISIT", "PD", "PK"), # Multiple domains
  run_analytics = TRUE,
  analytics_package = "gsm.kri", # Specify analytics package
  verbose = TRUE
)

# Generate study summary
summary_data <- summarize_longitudinal_study(study)
print(summary_data)
```

## Custom Base Dates

You can specify a custom starting date for your longitudinal study:

```{r longitudinal-custom-date}
# Use the lower-level function for more control
mappings <- ensure_core_mappings(c("AE", "LB", "VISIT"))

raw_data <- generate_study_snapshots(
  study_id = "CUSTOM-001",
  participants = 100,
  sites = 8,
  snapshots = 6,
  interval = "2 months", # Bi-monthly snapshots
  mappings = mappings,
  base_date = "2022-01-15", # Custom start date
  verbose = TRUE
)

# Create study object
config <- list(
  participants = 100,
  sites = 8,
  snapshots = 6,
  interval = "2 months",
  domains = c("AE", "LB", "VISIT")
)

study <- create_longitudinal_study_data(
  study_id = "CUSTOM-001",
  raw_data = raw_data,
  config = config
)
```

## Different Time Intervals

The package supports various time intervals for longitudinal studies:

```{r longitudinal-intervals}
# Weekly snapshots
weekly_study <- create_longitudinal_study(
  study_id = "WEEKLY-001",
  participants = 50,
  sites = 3,
  snapshots = 12,
  interval = "1 week", # Weekly intervals
  domains = c("AE", "VISIT"),
  verbose = TRUE
)

# Quarterly snapshots
quarterly_study <- create_longitudinal_study(
  study_id = "QUARTERLY-001",
  participants = 300,
  sites = 15,
  snapshots = 8,
  interval = "3 months", # Quarterly intervals
  domains = c("AE", "LB", "PD"),
  verbose = TRUE
)
```

## Quick Study Creation

For rapid prototyping, use the `quick_longitudinal_study()` function. Pass `include_pipeline = TRUE` to run both the analytics and reporting pipelines:

```{r quick-study}
# Create a study with sensible defaults — analytics and reporting both run by default
quick_study <- quick_longitudinal_study(
  study_name = "QUICK-PROTO-001",
  participants = 200,
  sites = 10,
  months_duration = 18, # 18-month study
  study_type = "standard",
  include_pipeline = TRUE, # runs analytics + reporting
  verbose = TRUE
)

# Analytics and reporting results are available on the study object
names(quick_study) # includes $analytics and $reporting
```

# Study Configuration Best Practices

## Domain Selection

Choose clinical domains based on your study objectives:

```{r domain-selection}
# Oncology study domains
oncology_domains <- c(
  "AE", "LB", "PD", "OverallResponse",
  "VISIT", "DATACHG", "QUERY"
)

# Safety study domains
safety_domains <- c(
  "AE", "LB", "VISIT",
  "Death", "STUDCOMP"
)

# PK/PD study domains
pkpd_domains <- c("PK", "PD", "LB", "VISIT", "AE")

# Create study with oncology focus
onc_study <- create_longitudinal_study(
  study_id = "ONCOLOGY-001",
  participants = 300,
  sites = 20,
  snapshots = 24,
  interval = "1 month",
  domains = oncology_domains,
  verbose = TRUE
)
```

## Analytics and Reporting Configuration

Configure the analytics package and specific workflows, and optionally customise the reporting package. The analytics pipeline returns `workr` results per snapshot; pass them directly to the reporting pipeline.

```{r analytics-config}
# Create study with specific analytics workflows
study <- create_longitudinal_study(
  study_id = "MONITORED-001",
  participants = 400,
  sites = 30,
  snapshots = 12,
  interval = "1 month",
  domains = c("AE", "LB", "VISIT", "QUERY"),
  run_analytics = TRUE,
  analytics_package = "gsm.kri",
  analytics_workflows = c("kri0001", "kri0002"), # Specific workflows
  run_reporting = TRUE,
  verbose = TRUE
)

# Access analytics: $results keyed by workflow name
if (!is.null(study$analytics)) {
  first_snap <- study$analytics[[1]]

  # All metric results for the first snapshot
  metric_names <- names(first_snap$results)

  # Access a specific metric and its data frames
  kri <- first_snap$results[[metric_names[1]]]
  kri$Analysis_Summary # summary-level data frame
  kri$Analysis_Flagged # flagged entities
  kri$Analysis_Analyzed # full analyzed data
}

# Access reporting: keyed by snapshot date
if (!is.null(study$reporting)) {
  names(study$reporting) # snapshot dates
  study$reporting[[1]] # reporting output for first snapshot
}
```

You can also run the pipelines separately on an existing study object:

```{r pipeline-stepwise}
# Build raw data only
study <- create_longitudinal_study(
  study_id = "STEPWISE-001",
  participants = 200,
  sites = 10,
  snapshots = 6,
  interval = "1 month",
  domains = c("AE", "LB"),
  run_analytics = FALSE # skip pipelines for now
)

# Run analytics when ready
study <- run_longitudinal_analytics(study, verbose = TRUE)

# Run reporting on top of analytics
study <- run_longitudinal_reporting(study, verbose = TRUE)
```

# Multiple Studies Generation

When you need to generate data for multiple studies simultaneously, the `create_multiple_longitudinal_studies()` function provides an efficient approach. This is particularly useful for:

- Comparing different study designs
- Generating test data for multiple protocols
- Batch processing for portfolio analysis
- Creating matched cohorts with different characteristics

## Basic Multiple Studies Setup

Create multiple studies with shared configuration:

```{r multiple-basic}
# Generate three studies with identical configuration
studies <- create_multiple_longitudinal_studies(
  study_names = c("PHASE2-001", "PHASE2-002", "PHASE2-003"),
  participants = 150,
  sites = 12,
  snapshots = 6,
  interval = "1 month",
  domains = c("AE", "LB", "VISIT", "PD"),
  run_analytics = TRUE,
  run_reporting = FALSE,
  verbose = TRUE
)

# Examine the collection
print(studies) # Shows summary of all studies
names(studies) # Study names

# Access individual studies
study_1 <- studies[["PHASE2-001"]]
summary(study_1)
```

## Per-Study Configuration

Customize individual studies while maintaining shared defaults:

```{r multiple-custom}
# Create studies with different characteristics
studies <- create_multiple_longitudinal_studies(
  study_names = c("SMALL-PHASE2", "LARGE-PHASE3", "SAFETY-RUN"),
  participants = c(80, 400, 50), # Different participant counts
  sites = c(8, 25, 3), # Different site counts
  snapshots = c(4, 12, 8), # Different durations
  interval = "1 month", # Shared interval
  domains = c("AE", "LB", "VISIT"), # Base domains (can be overridden)
  study_configs = list(
    "SMALL-PHASE2" = list(
      domains = c("AE", "LB", "VISIT"),
      interval = "2 weeks",
      outlier_intensity = 0.8
    ),
    "LARGE-PHASE3" = list(
      domains = c("AE", "LB", "VISIT", "PD", "PK", "QUERY"),
      analytics_package = "gsm.kri",
      analytics_workflows = c("kri0001", "kri0002", "kri0005")
    ),
    "SAFETY-RUN" = list(
      domains = c("AE", "LB", "Death"),
      outlier_intensity = 2.0,
      interval = "1 week"
    )
  ),
  run_analytics = TRUE,
  verbose = TRUE
)

# Each study has different characteristics
summary(studies)
```

## Parallel Processing

For large batches, enable parallel processing to speed up generation:

```{r multiple-parallel}
# Generate studies in parallel (requires parallel package)
large_batch <- create_multiple_longitudinal_studies(
  study_names = paste0("BATCH-", sprintf("%03d", 1:8)),
  participants = 200,
  sites = 15,
  snapshots = 6,
  domains = c("AE", "LB", "VISIT"),
  parallel = TRUE, # Enable parallel processing
  run_analytics = TRUE,
  verbose = TRUE
)

# Check results
print(large_batch)
```

## Export Multiple Studies

Automatically export all studies to disk during generation:

```{r multiple-export}
# Generate and automatically export
studies <- create_multiple_longitudinal_studies(
  study_names = c("EXPORT-001", "EXPORT-002"),
  participants = 100,
  sites = 8,
  snapshots = 4,
  domains = c("AE", "LB"),
  run_analytics = TRUE,
  run_reporting = TRUE,
  export_studies = TRUE, # Auto-export to disk
  export_dir = "./studies", # Export directory
  verbose = TRUE
)

# Or export afterwards
export_multiple_studies(
  studies = studies,
  output_dir = "./exported_studies",
  save_rds = TRUE,
  verbose = TRUE
)
```

## Working with Multiple Studies Collections

The returned object has special methods for easy inspection:

```{r multiple-methods}
# Print method shows overview
print(studies)

# Summary method provides detailed statistics
study_summary <- summary(studies)
print(study_summary)

# Access individual studies
individual_study <- studies[["EXPORT-001"]]

# Loop through all studies
for (study_name in names(studies)) {
  study <- studies[[study_name]]
  cat("Study:", study_name, "\n")
  cat("  Participants:", study$config$participants, "\n")
  cat("  Analytics:", !is.null(study$analytics), "\n")
}
```

## Portfolio-Level Analysis

Use multiple studies for comparative analysis:

```{r multiple-portfolio}
# Generate studies with different safety profiles
safety_portfolio <- create_multiple_longitudinal_studies(
  study_names = c("LOW-RISK", "MEDIUM-RISK", "HIGH-RISK"),
  participants = 200,
  sites = 12,
  snapshots = 8,
  domains = c("AE", "LB", "Death", "STUDCOMP"),
  study_configs = list(
    "LOW-RISK" = list(outlier_intensity = 0.5),
    "MEDIUM-RISK" = list(outlier_intensity = 1.0),
    "HIGH-RISK" = list(outlier_intensity = 2.5)
  ),
  run_analytics = TRUE,
  verbose = TRUE
)

# Compare AE rates across studies
for (study_name in names(safety_portfolio)) {
  study <- safety_portfolio[[study_name]]

  # Get final snapshot AE data
  final_snapshot <- study$raw_data[[length(study$raw_data)]]
  ae_count <- if ("Raw_AE" %in% names(final_snapshot)) nrow(final_snapshot$Raw_AE) else 0

  cat(sprintf("%s: %d total AEs\n", study_name, ae_count))
}
```

# Data Examination and Validation

## Exploring Generated Data

```{r data-exploration}
# After generating a study, examine the data structure
study <- create_longitudinal_study(
  study_id = "EXPLORE-001",
  participants = 50,
  sites = 3,
  snapshots = 3,
  interval = "1 month",
  domains = c("AE", "LB"),
  verbose = TRUE,
  run_analytics = FALSE
)

# Check data for each snapshot
for (i in 1:length(study$raw_data)) {
  snapshot_name <- names(study$raw_data)[i]
  snapshot_data <- study$raw_data[[i]]

  cat("Snapshot:", snapshot_name, "\n")
  cat("Datasets:", paste(names(snapshot_data), collapse = ", "), "\n")

  # Check sample sizes
  if ("Raw_SUBJ" %in% names(snapshot_data)) {
    cat("Subjects:", nrow(snapshot_data$Raw_SUBJ), "\n")
  }
  if ("Raw_AE" %in% names(snapshot_data)) {
    cat("AE Records:", nrow(snapshot_data$Raw_AE), "\n")
  }
  cat("\n")
}
```

## Data Quality Checks

```{r data-quality}
# Function to validate study data
validate_study_data <- function(study) {
  validation_results <- list()

  for (snapshot_name in names(study$raw_data)) {
    snapshot <- study$raw_data[[snapshot_name]]

    # Check required datasets
    required_datasets <- c("Raw_STUDY", "Raw_SITE", "Raw_SUBJ", "Raw_ENROLL")
    missing_required <- setdiff(required_datasets, names(snapshot))

    validation_results[[snapshot_name]] <- list(
      datasets_present = names(snapshot),
      missing_required = missing_required,
      total_subjects = if ("Raw_SUBJ" %in% names(snapshot)) nrow(snapshot$Raw_SUBJ) else 0,
      total_sites = if ("Raw_SITE" %in% names(snapshot)) nrow(snapshot$Raw_SITE) else 0
    )
  }

  return(validation_results)
}

# Validate a study
validation <- validate_study_data(study)
str(validation)
```

# Exporting Study Data

Once you have a complete study object — with raw data, analytics, and (optionally) reporting — you can write everything to disk with `export_study_data()`. The function creates a structured folder hierarchy under a root directory:

```
<output_dir>/<study_id>/
  <snapshot_date>/
    raw/          # Raw_*.csv
    mapped/       # Mapped_*.csv  (present when analytics ran)
    analytics/    # <metric>_<table>.csv  (present when analytics ran)
    reporting/    # Reporting_*.csv  (present when reporting ran)
```

Folders are only created when the corresponding data exists for a snapshot.

## Basic Export

```{r export-basic}
# Generate a small study to export
study <- create_longitudinal_study(
  study_id = "EXPORT-001",
  participants = 100,
  sites = 5,
  snapshots = 3,
  interval = "1 month",
  domains = c("AE", "LB"),
  run_analytics = TRUE,
  run_reporting = TRUE,
  verbose = FALSE
)

# Export to a temporary directory — returns the path to the study folder invisibly
study_path <- export_study_data(
  study      = study,
  output_dir = tempdir(),
  verbose    = TRUE # prints per-snapshot progress
)

# Confirm the folder structure
list.dirs(study_path, full.names = FALSE, recursive = TRUE)
```

## Controlling the Output Location

```{r export-location}
# Write to a specific project folder
study_path <- export_study_data(
  study        = study,
  output_dir   = "~/my_studies", # root folder
  study_folder = "EXPORT-001-v2", # override the auto-generated folder name
  overwrite    = TRUE # allow writing into an existing folder
)
```

## Preserving Non-CSV Objects

The analytics pipeline stores workflow lists, summaries, and other objects that cannot be flattened to CSV. Set `save_rds = TRUE` to write a companion `analytics_full.rds` file per snapshot:

```{r export-rds}
study_path <- export_study_data(
  study      = study,
  output_dir = tempdir(),
  overwrite  = TRUE,
  save_rds   = TRUE # also writes analytics_full.rds per snapshot
)

# Reload the full analytics object for a specific snapshot
analytics_snap <- readRDS(file.path(
  study_path, names(study$raw_data)[1],
  "analytics_full.rds"
))
names(analytics_snap) # $results, $mapped, $lWorkflow, $summary
```

## Inspecting the Exported Files

```{r export-inspect}
# List all CSV files written across all snapshots
all_csvs <- list.files(study_path, pattern = "\\.csv$", recursive = TRUE)
cat(length(all_csvs), "CSV files written\n")

# Raw data lives in <snapshot>/raw/
raw_csvs <- all_csvs[grepl("/raw/", all_csvs)]
cat("Raw CSVs:", paste(basename(raw_csvs), collapse = ", "), "\n")

# Read a specific snapshot's adverse event data back in
snap_date <- names(study$raw_data)[1]
ae_path <- file.path(study_path, snap_date, "raw", "Raw_AE.csv")
ae_df <- read.csv(ae_path)
nrow(ae_df)
```

# Conclusion

The `gsm.datasim` package provides flexible tools for creating both single snapshot and longitudinal clinical studies. Key takeaways:

- Use `create_study_config()` and `generate_study_data()` for single snapshots
- Use `create_longitudinal_study()` for multi-snapshot studies
- Configure domains based on your study type and monitoring needs
- Leverage the analytics pipeline for automated quality monitoring
- Use `quick_longitudinal_study()` for rapid prototyping
- Use `export_study_data()` to write completed studies to a structured folder hierarchy

The modular design allows you to start simple and add complexity as needed, making it suitable for both exploratory work and production data generation scenarios.
