---
title: "Domain Registry"
author: "[gsm.datasim](https://gilead-public.github.io/gsm.datasim) Example"
description: "How to inspect, use, and extend the Domain Registry for custom data generation."
index: 3
date: "`r format(Sys.time(), '%B %d, %Y %H:%M:%S %Z')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(gsm.datasim)
```

## Overview

The Domain Registry is the extensible system that replaces the legacy
`switch()` block for domain generation. It defines, per domain, exactly
how data should be generated: what counts to use, what arguments to pass,
and which generator function to call.

Migrating a domain to the registry makes it:

- **Self-describing** — all logic lives in one place per domain
- **Independently testable**
- **Overridable** without touching core generation code

## Inspecting the Registry

```{r inspect}
registry <- get_domain_registry()

cat("Registry-backed domains:", paste(names(registry), collapse = ", "), "\n")

# Each entry has four contract fields
cat("Raw_AE entry fields:", paste(names(registry$Raw_AE), collapse = ", "), "\n")
# > "dataset"  "required_inputs"  "count_fn"  "generate_fn"
```

## The Entry Contract

Every valid entry must have exactly these four fields:

| Field             | Type                         | Description                                               |
|-------------------|------------------------------|-----------------------------------------------------------|
| `$dataset`        | `character(1)`               | The `Raw_*` name                                          |
| `$required_inputs`| `character(n)`               | Context keys the entry reads                              |
| `$count_fn`       | `function(counts, snapshot_idx)` | Picks the right count for this snapshot              |
| `$generate_fn`    | `function(context)`          | All generation logic; returns a `data.frame`              |

```{r entry-contract}
ae_entry <- registry$Raw_AE

cat("dataset        :", ae_entry$dataset, "\n")
cat("required_inputs:", paste(ae_entry$required_inputs, collapse = ", "), "\n")

# count_fn body
print(body(ae_entry$count_fn))

# generate_fn body (first 10 lines)
bdy <- deparse(body(ae_entry$generate_fn))
cat(paste(head(bdy, 10), collapse = "\n"), "\n  ...\n")
```

## How the Registry Fits the Pipeline

Inside `run_domain_generation_loop()` the per-domain dispatch is:

```r
migrated_data <- generate_domain_from_registry(data_type, context)
if (!is.null(migrated_data)) {
  data[[data_type]] <- migrated_data
  next   # skip the legacy dispatcher
}
# ... dispatch_legacy_domain_generator() for domains not yet in the registry
```

The `context` list passed to every `generate_fn` contains:

```
data, previous_data, combined_specs,
n, start_date, end_date,
snapshot_idx, snapshot_count, snapshot_width, study_id
```

You can verify the registry path end-to-end by generating a small study:

```{r pipeline-demo}
config <- create_study_config(
  study_id          = "REG-DEMO-001",
  participant_count = 50,
  site_count        = 5
) |>
  set_temporal_config(
    start_date     = "2023-01-01",
    snapshot_count = 3,
    snapshot_width = "months"
  ) |>
  add_dataset_config("Raw_AE",    enabled = TRUE) |>
  add_dataset_config("Raw_LB",    enabled = TRUE) |>
  add_dataset_config("Raw_VISIT", enabled = TRUE)

raw_data <- generate_raw_data_from_config(config, verbose = TRUE)

cat("Datasets in snapshot 1:", paste(names(raw_data[[1]]), collapse = ", "), "\n")
cat("Raw_AE rows  :", nrow(raw_data[[1]]$Raw_AE),   "\n")
cat("Raw_LB rows  :", nrow(raw_data[[1]]$Raw_LB),   "\n")
cat("Raw_VISIT rows:", nrow(raw_data[[1]]$Raw_VISIT), "\n")
```

## Adding a New Domain to the Registry

To migrate a domain from the legacy dispatcher:

1. Write a well-typed entry list
2. Register it via `raw_data_generator()` override or by editing
   `get_domain_registry()` directly
3. Remove the corresponding `case` from `dispatch_legacy_domain_generator()`

```{r add-domain}
# Example: migrating a hypothetical Raw_Biomarker domain
new_entry <- list(
  dataset         = "Raw_Biomarker",
  required_inputs = c("data", "previous_data", "combined_specs", "n", "start_date"),
  count_fn        = function(counts, snapshot_idx) counts$subject_count[snapshot_idx],
  generate_fn     = function(context) {
    spec          <- context$combined_specs
    data          <- context$data
    previous_data <- context$previous_data
    n             <- context$n

    if ("Raw_Biomarker" %in% names(previous_data)) {
      dataset          <- previous_data$Raw_Biomarker
      previous_row_num <- nrow(dataset)
    } else {
      dataset          <- NULL
      previous_row_num <- 0
    }

    n_new <- n - previous_row_num
    if (n_new <= 0) return(dataset)

    args <- list(
      subjid  = list(n_new, external_subjid = data$Raw_SUBJ$subjid),
      default = list(n_new, context$start_date)
    )

    as.data.frame(add_new_var_data(dataset, spec$Raw_Biomarker, args, spec$Raw_Biomarker))
  }
)
```

## Using the Config Helper Path

Equivalent generation using the explicit config helper:

```{r config-path}
config <- create_study_config(
  study_id          = "REGISTRY-EXAMPLE-001",
  participant_count = 80,
  site_count        = 8
) |>
  set_temporal_config(
    start_date     = "2012-01-01",
    snapshot_count = 2,
    snapshot_width = "months"
  )

for (ds in c("Raw_STUDY", "Raw_SITE", "Raw_SUBJ", "Raw_ENROLL",
             "Raw_VISIT", "Raw_AE", "Raw_LB")) {
  config <- add_dataset_config(config, ds, enabled = TRUE)
}

raw_data            <- generate_study_data(config)
raw_data_via_config <- generate_raw_data_from_config(config)

cat("Snapshots:", length(raw_data), "\n")
cat("Snapshot keys:", paste(names(raw_data), collapse = ", "), "\n")

snapshot_1 <- raw_data[[1]]
cat("Datasets in snapshot 1:", paste(names(snapshot_1), collapse = ", "), "\n")
cat("Raw_AE rows:", nrow(snapshot_1$Raw_AE), "\n")
cat("Raw_LB rows:", nrow(snapshot_1$Raw_LB), "\n")

# Wrap into a study object for helper functions
study <- create_longitudinal_study_data(
  study_id = "REGISTRY-EXAMPLE-001",
  raw_data = raw_data,
  config   = list(
    participants = 80,
    sites        = 8,
    snapshots    = 2,
    interval     = "1 month",
    domains      = c("AE", "LB", "VISIT"),
    study_type   = "standard"
  )
)

summarize_longitudinal_study(study)
```
