---
title: "Generating Multiple Studies"
author: "[gsm.datasim](https://gilead-biostats.github.io/gsm.datasim) Example"
description: "Shows three approaches to generating a portfolio of studies: identical configs, per-study vectors, and the study_portfolio() wrapper for variant-based configs."
index: 4
date: "`r format(Sys.time(), '%B %d, %Y %H:%M:%S %Z')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(gsm.datasim)
```

## Overview

`gsm.datasim` provides two functions for generating multiple studies in a
single call:

| Function | Best for |
|---|---|
| `create_multiple_longitudinal_studies()` | Full control; vectorised params + `study_configs` overrides |
| `study_portfolio()` | Variant-first design — define only what differs per study |

Both return a `multiple_longitudinal_studies` object (a named list of
`longitudinal_study` objects) with `print()` and `summary()` methods.

---

## 1. Identical configuration across all studies

When every study shares the same design, pass a single value for each
parameter and a vector of names.

```{r identical}
studies <- create_multiple_longitudinal_studies(
  study_names   = c("TRIAL-001", "TRIAL-002", "TRIAL-003"),
  participants  = 150,
  sites         = 12,
  snapshots     = 6,
  interval      = "1 month",
  domains       = c("AE", "LB", "VISIT", "PD"),
  run_analytics = FALSE,
  verbose       = TRUE
)

print(studies)          # collection-level summary
summary(studies)        # aggregate statistics

# Access a single study
trial_1 <- studies[["TRIAL-001"]]
names(trial_1$raw_data)                         # snapshot dates
nrow(trial_1$raw_data[[1]][["Raw_AE"]])         # AE rows in snapshot 1
```

---

## 2. Per-study parameters via vectors and `study_configs`

Pass parallel vectors for scalar parameters and use `study_configs` for
anything that can't be vectorised (e.g. `domains`).

```{r per-study-vectors}
studies <- create_multiple_longitudinal_studies(
  study_names  = c("PHASE2-SMALL", "PHASE3-LARGE", "SAFETY-RUN"),
  # Scalar params as vectors -- one value per study
  participants = c(80,  400, 50),
  sites        = c(8,   25,  3),
  snapshots    = c(4,   12,  8),
  interval     = "1 month",              # shared
  domains      = c("AE", "LB", "VISIT"), # default; overridden below
  # Non-scalar overrides per study
  study_configs = list(
    "PHASE2-SMALL" = list(
      domains           = c("AE", "LB"),
      outlier_intensity = 0.8
    ),
    "PHASE3-LARGE" = list(
      domains           = c("AE", "LB", "VISIT", "PD", "PK"),
      outlier_intensity = 1.0
    ),
    "SAFETY-RUN" = list(
      domains           = c("AE", "LB", "VISIT"),
      outlier_intensity = 2.0   # amplify outliers for safety monitoring testing
    )
  ),
  run_analytics = FALSE,
  verbose       = TRUE
)

# Confirm each study got its own config
for (nm in names(studies)) {
  cat(nm, "-- participants:", studies[[nm]]$config$participants,
      "| domains:", paste(studies[[nm]]$config$domains, collapse = ", "), "\n")
}
```

---

## 3. `study_portfolio()` -- variant-first interface

Define a single named list of variants.  Each entry contains only what
differs from the shared defaults; everything else falls back automatically.

```{r portfolio}
studies <- study_portfolio(
  variants = list(
    "PHASE2-SMALL" = list(
      participants      = 80,
      sites             = 8,
      snapshots         = 4,
      domains           = c("AE", "LB"),
      outlier_intensity = 0.8
    ),
    "PHASE3-LARGE" = list(
      participants = 400,
      sites        = 25,
      snapshots    = 12,
      domains      = c("AE", "LB", "VISIT", "PD", "PK")
    ),
    "SAFETY-RUN" = list(
      participants      = 50,
      sites             = 3,
      snapshots         = 8,
      outlier_intensity = 2.0
    )
  ),
  # Shared defaults -- used when a variant does not specify its own value
  participants  = 100,
  sites         = 10,
  snapshots     = 6,
  interval      = "1 month",
  domains       = c("AE", "LB", "VISIT"),
  run_analytics = FALSE,
  verbose       = TRUE
)

print(studies)
```

---

## 4. Inspecting and comparing studies

```{r inspect}
# Snapshot dates for each study
lapply(studies, function(s) names(s$raw_data))

# Row counts for Raw_AE at the final snapshot, across all studies
sapply(studies, function(s) {
  last_snap <- tail(s$raw_data, 1)[[1]]
  if ("Raw_AE" %in% names(last_snap)) nrow(last_snap$Raw_AE) else NA_integer_
})

# Domains present in each study
lapply(studies, get_available_domains)

# Timeline of a single domain across snapshots for one study
ae_timeline <- get_domain_timeline(studies[["PHASE3-LARGE"]], "Raw_AE")
sapply(ae_timeline, nrow)   # row count per snapshot
```

---

## 5. Running analytics and reporting

Pass `run_analytics = TRUE` (and optionally `run_reporting = TRUE`) to run
the full pipeline for every study in one call.

```{r with-analytics}
studies <- study_portfolio(
  variants = list(
    "PHASE2-SMALL" = list(participants = 80,  sites = 8,  snapshots = 4,
                          domains = c("AE", "LB")),
    "PHASE3-LARGE" = list(participants = 400, sites = 25, snapshots = 12,
                          domains = c("AE", "LB", "VISIT", "PD"))
  ),
  participants  = 100,
  sites         = 10,
  snapshots     = 6,
  interval      = "1 month",
  domains       = c("AE", "LB", "VISIT"),
  run_analytics = TRUE,
  run_reporting = FALSE,
  verbose       = TRUE
)

# Analytics results are nested: study -> snapshot -> results
has_analytics <- sapply(studies, function(s) !is.null(s$analytics))
cat("Analytics completed:", sum(has_analytics), "of", length(studies), "studies\n")

# Drill into one study's latest snapshot analytics
latest  <- tail(names(studies[["PHASE3-LARGE"]]$analytics), 1)
results <- studies[["PHASE3-LARGE"]]$analytics[[latest]]$results
cat("Result tables in latest snapshot:", paste(names(results), collapse = ", "), "\n")
```

---

## 6. Exporting all studies

```{r export}
# Export to separate sub-folders under ./output/
export_multiple_studies(
  studies    = studies,
  output_dir = "./output",
  overwrite  = TRUE,
  verbose    = TRUE
)

# Or generate and export in one step via create_multiple_longitudinal_studies()
studies <- create_multiple_longitudinal_studies(
  study_names    = c("TRIAL-001", "TRIAL-002"),
  participants   = 150,
  sites          = 12,
  snapshots      = 6,
  domains        = c("AE", "LB", "VISIT"),
  run_analytics  = FALSE,
  export_studies = TRUE,
  export_dir     = "./output",
  verbose        = TRUE
)
```
