---
title: "Test Data Explorer"
author: "[gsm.datasim](https://gilead-public.github.io/gsm.datasim) Example"
description: "Browse the shared gsm test data bundle: what is in it, how big each table is, and how it changes across snapshots."
index: 5
date: "`r format(Sys.time(), '%B %d, %Y %H:%M:%S %Z')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  message  = FALSE,
  warning  = FALSE
)
library(gsm.datasim)
```

## The shared test data bundle

The gsm ecosystem develops and qualifies against one shared set of simulated
study data. It used to ship inside `gsm.core` as `lSource` and the
`reporting*` / `analytics*` objects, so every data refresh needed a `gsm.core`
release. It is now built here by `build_testdata_bundle()` from the checked-in
[`bundle-config.yaml`](https://github.com/Gilead-Public/gsm.datasim/blob/dev/inst/testdata/bundle-config.yaml)
and written as a versioned bundle with a `manifest.json` describing exactly how
it was produced.

This page renders the explorer for the bundle found in `GSM_TESTDATA_DIR` when
that variable is set (the published bundle on CI), and otherwise builds a
reduced bundle on the fly so the page works anywhere the pipeline packages are
installed.

```{r bundle}
bundle_dir <- Sys.getenv("GSM_TESTDATA_DIR", unset = "")

if (nzchar(bundle_dir)) {
  bundle <- read_testdata_bundle(bundle_dir)
} else {
  cfg <- read_testdata_config(
    bundle_version = "0.0.0", # local preview, not a published version
    participants = 200,
    sites = 20
  )
  bundle <- build_testdata_bundle(cfg, output_dir = tempdir(), overwrite = TRUE)
}
bundle
```

## Explore

Select a table in the tree to page through its rows or inspect its schema.
The **Overview** shows how each table grows across snapshots and a few
distributions from the latest raw layer. The latest `raw/` layer (the
`lSource` equivalent) embeds up to 5,000 rows per table and other tables up
to 1,000, so the page stays small; the summaries always cover the whole table
and the full bundle is available for download.

```{r explorer, out.width="100%"}
explore_testdata(bundle)
```

## What is in the bundle

```{r inventory}
s <- summarize_testdata_bundle(bundle)
knitr::kable(
  aggregate(cbind(tables = 1, rows = rows) ~ snapshot + layer, data = s$tables, FUN = sum),
  caption = "Tables and rows per snapshot and layer"
)
```

## Reproducing the bundle

```{r reproduce, eval=FALSE}
# the production configuration (1000 participants, 150 sites, 3 snapshots)
cfg <- read_testdata_config()
bundle <- build_testdata_bundle(cfg, output_dir = "~/gsm-testdata")

# the manifest records the seed, parameters, package versions and a checksum per file
bundle$manifest$package_versions
head(bundle$manifest$files)
```
