---
title: "Workflow-Driven Data Generation"
author: "[gsm.datasim](https://gilead-public.github.io/gsm.datasim) Example"
description: "Generate simulated raw data directly from an lWorkflows spec — the same structure returned by workr::MakeWorkflowList(). Covers single-snapshot, longitudinal, domain subsetting, and custom/unknown domains."
index: 4
date: "`r format(Sys.time(), '%B %d, %Y %H:%M:%S %Z')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
library(gsm.datasim)
```

## Overview

`generate_data_from_workflows()` accepts an `lWorkflows` list — the same
structure returned by `workr::MakeWorkflowList()` — and produces `Raw_*`
data frames for every domain defined in those specs.

Domains with dedicated generators (`Raw_AE`, `Raw_SUBJ`, etc.) use curated
logic. Unknown domains fall back to **type-based column generation** driven
by the spec metadata, so any custom workflow produces a usable data frame
without extra code.

## Load Workflows

```{r load-workflows}
lWorkflows <- workr::MakeWorkflowList(
  strPath    = "workflow/1_mappings",
  strPackage = "gsm.mapping"  # or any package with workr-compatible workflow specs
)
```

## Single-Snapshot Generation

Generate raw data for all domains in the spec in one call.

```{r single-snapshot}
raw_data <- generate_data_from_workflows(
  lWorkflows     = lWorkflows,
  n_participants = 200,
  n_sites        = 20,
  study_id       = "WF-DEMO-001",
  start_date     = "2012-01-01",
  end_date       = "2012-12-31"
)

cat("Generated domains:\n")
for (nm in names(raw_data)) {
  cat(sprintf("  %-25s %d rows x %d cols\n", nm, nrow(raw_data[[nm]]), ncol(raw_data[[nm]])))
}
```

## Custom Domain Counts

Override the default row-count formula for specific domains.

```{r custom-counts}
raw_data_custom <- generate_data_from_workflows(
  lWorkflows     = lWorkflows,
  n_participants = 100,
  domain_counts  = list(Raw_AE = 500, Raw_PD = 150)
)
```

## Longitudinal Multi-Snapshot Generation

Generate 6 monthly snapshots. Row counts ramp up via `count_gen()` so early
snapshots have fewer participants than later ones. Each snapshot's data is
cumulative — rows from snapshot N are preserved in snapshot N+1.

```{r longitudinal}
snapshots <- generate_data_from_workflows(
  lWorkflows     = lWorkflows,
  n_participants = 200,
  n_sites        = 20,
  study_id       = "WF-LONG-001",
  start_date     = "2012-01-01",
  snapshot_count = 6,
  snapshot_width = "months"
)

cat("Longitudinal snapshots:\n")
for (snap_name in names(snapshots)) {
  snap   <- snapshots[[snap_name]]
  n_subj <- if ("Raw_SUBJ" %in% names(snap)) nrow(snap$Raw_SUBJ) else NA
  cat(sprintf("  %s: %d domains, Raw_SUBJ = %s rows\n",
              snap_name, length(snap), n_subj))
}
```

## Selecting a Domain Subset

Pass `desired_domains` to generate only the domains you need.

```{r subset}
raw_data_subset <- generate_data_from_workflows(
  lWorkflows      = lWorkflows,
  n_participants  = 50,
  desired_domains = c("Raw_SUBJ", "Raw_AE", "Raw_SITE")
)
```

## Custom Workflows and Unknown Domains

For domains without a dedicated generator, columns are produced based on
their spec `type` or name pattern. This means any custom workflow produces
a usable data frame without extra code.

```{r custom-domain}
custom_workflows <- list(
  custom_disease = list(
    meta  = list(Description = "Custom disease assessment"),
    spec  = list(
      Raw_DISEASE = list(
        subjid      = list(required = TRUE),
        assess_dt   = list(type = "date"),
        score_val   = list(type = "numeric"),
        category    = list(type = "character"),
        resolved_yn = list(required = TRUE)
      )
    ),
    steps = list()
  )
)

raw_custom <- generate_data_from_workflows(
  lWorkflows     = custom_workflows,
  n_participants = 30,
  n_sites        = 5,
  study_id       = "CUSTOM-001"
)

cat("Custom domain columns:\n")
str(raw_custom$Raw_DISEASE)
```

## On-the-Fly Column Overrides

`column_overrides` lets you add new columns or replace existing ones in any
generated domain without modifying workflow specs or generator functions.
Overrides are applied **after** the domain is generated, so all curated
structure (e.g. `Raw_LB`'s test panel / subject-visit cross-join) is preserved.

Each column value can be a function, a vector to sample from, or a scalar.

### Add a new column with a custom distribution

If a preexisting workflow spec references a column that has no named generator
(e.g. `score_val` in an LB workflow), `gsm.datasim` auto-fills it via
type inference. Use `column_overrides` when you need a specific distribution
instead of the generic fallback.

```{r overrides-function}
raw_lb_scored <- generate_data_from_workflows(
  lWorkflows     = lWorkflows,
  n_participants = 200,
  column_overrides = list(
    Raw_LB = list(
      score_val = function(n) round(runif(n, 0, 10), 1)
    )
  )
)

summary(raw_lb_scored$Raw_LB$score_val)
```

### Sample from a fixed set of values

Pass a vector and each row is drawn with replacement.

```{r overrides-vector}
raw_lb_units <- generate_data_from_workflows(
  lWorkflows     = lWorkflows,
  n_participants = 200,
  column_overrides = list(
    Raw_LB = list(
      lbstresu = c("mg/dL", "mmol/L", "g/L")
    )
  )
)

table(raw_lb_units$Raw_LB$lbstresu)
```

### Broadcast a scalar to every row

A length-1 value is repeated to fill all rows.

```{r overrides-scalar}
raw_lb_cat <- generate_data_from_workflows(
  lWorkflows     = lWorkflows,
  n_participants = 200,
  column_overrides = list(
    Raw_LB = list(lbcat = "CHEMISTRY")
  )
)

unique(raw_lb_cat$Raw_LB$lbcat)
```

### Derive a column from the generated data frame

Use `function(n, df)` (two parameters) to access the fully-generated domain
data frame and compute a column from existing columns.

```{r overrides-derive}
raw_lb_derived <- generate_data_from_workflows(
  lWorkflows     = lWorkflows,
  n_participants = 200,
  column_overrides = list(
    Raw_LB = list(
      lbstnrhi  = function(n, df) round(df$lbstresn * 1.2, 2),
      visit_flag = function(n, df) ifelse(df$visnam == "SCREENING", "S", "F")
    )
  )
)

head(raw_lb_derived$Raw_LB[, c("lbstresn", "lbstnrhi", "visnam", "visit_flag")])
```

### Combine overrides across multiple domains

```{r overrides-multi}
raw_multi <- generate_data_from_workflows(
  lWorkflows     = lWorkflows,
  n_participants = 200,
  column_overrides = list(
    Raw_LB = list(
      lbcat     = "CHEMISTRY",
      score_val = function(n) round(runif(n, 0, 10), 1)
    ),
    Raw_AE = list(
      severity_score = function(n) sample(1:5, n, replace = TRUE)
    )
  )
)
```

### Overrides in multi-snapshot mode

`column_overrides` are applied on every snapshot automatically.

```{r overrides-longitudinal}
snapshots_scored <- generate_data_from_workflows(
  lWorkflows     = lWorkflows,
  n_participants = 200,
  snapshot_count = 6,
  snapshot_width = "months",
  column_overrides = list(
    Raw_LB = list(
      score_val = function(n) round(runif(n, 0, 10), 1)
    )
  )
)

# score_val is present in every snapshot
sapply(snapshots_scored, function(s) !is.null(s$Raw_LB$score_val))
```
