---
title: "gsm.datasim Demo"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
library(gsm.datasim)
```

## Data Generation Only

If you only need simulated raw data — no analytics, no reporting — use
`create_longitudinal_study()` with both pipelines disabled. This is the
fastest way to get `Raw_*` data frames across multiple snapshots.

```{r data-only}
data_only <- create_longitudinal_study(
  study_id      = "DATA-001",
  participants  = 200,
  sites         = 15,
  snapshots     = 6,
  interval      = "1 month",
  domains       = c("AE", "LB", "VISIT", "QUERY", "SUBJ"),
  run_analytics = FALSE,
  run_reporting = FALSE,
  verbose       = TRUE
)

names(data_only)            # $study_id, $config, $raw_data
names(data_only$raw_data)   # one entry per snapshot date

snap  <- data_only$raw_data[[1]]
snap4 <- data_only$raw_data[[4]]
names(snap)
nrow(snap$Raw_SUBJ)
nrow(snap4$Raw_SUBJ)
nrow(snap$Raw_AE)
nrow(snap4$Raw_AE)
```

## Quick Start

`quick_longitudinal_study()` is the batteries-included entry point. It
generates multi-snapshot raw data with the domain list determined by
`study_type`, then runs the gsm.kri analytics pipeline and the
gsm.reporting pipeline — all in one call.

```{r quick-start}
study <- quick_longitudinal_study(
  study_name       = "DEMO-001",
  participants     = 200,
  sites            = 15,
  months_duration  = 6,
  study_type       = "standard",
  include_pipeline = TRUE
)
```

## Outlier Intensity

Use `outlier_intensity` to increase or decrease outlier-like values that
drive downstream flags in the analysis layer.

```{r outlier}
# Baseline
baseline_study <- quick_longitudinal_study(
  study_name       = "DEMO-OUTLIER-BASELINE",
  participants     = 200,
  sites            = 15,
  months_duration  = 6,
  study_type       = "standard",
  outlier_intensity = 1
)

# More aggressive outlier generation
high_outlier_study <- quick_longitudinal_study(
  study_name        = "DEMO-OUTLIER-HIGH",
  participants      = 200,
  sites             = 15,
  months_duration   = 6,
  study_type        = "standard",
  outlier_intensity = 2.5
)

# Also available on create_longitudinal_study()
custom_outlier_study <- create_longitudinal_study(
  study_id          = "DEMO-OUTLIER-CUSTOM",
  participants      = 150,
  sites             = 10,
  snapshots         = 4,
  interval          = "1 month",
  domains           = c("AE", "LB", "VISIT", "QUERY"),
  run_analytics     = TRUE,
  run_reporting     = FALSE,
  outlier_intensity = 2
)
```

## The Study Object

The returned list has five top-level slots.

```{r study-object}
names(study)
# $study_id   — character study identifier
# $config     — study parameters (participants, sites, domains, intervals, …)
# $raw_data   — named list of snapshot data frames, one entry per snapshot date
# $analytics  — gsm.kri pipeline output, named by snapshot date
# $reporting  — gsm.reporting pipeline output, named by snapshot date
```

## Raw Data

Snapshots are keyed by date (`YYYY-MM-DD`). Each snapshot is a named list
of data frames (`Raw_SUBJ`, `Raw_AE`, `Raw_LB`, …).

```{r raw-data}
names(study$raw_data)

snap_1 <- get_snapshot_data(study, 1)
names(snap_1)

cat("Subjects in snapshot 1:", nrow(snap_1$Raw_SUBJ), "\n")
cat("AE rows in snapshot 1: ", nrow(snap_1$Raw_AE),   "\n")

# Track a single domain across all snapshots
ae_timeline <- get_domain_timeline(study, "AE")
ae_counts   <- sapply(ae_timeline, nrow)
print(ae_counts)

get_available_domains(study)
```

## Analytics Pipeline Results

`study$analytics` is a named list (one entry per snapshot). Each snapshot
slot contains:

- `$results`   — named list of metric results (`Analysis_kri0001`, …)
- `$mapped`    — mapped data frames (`Mapped_SUBJ`, `Mapped_AE`, …)
- `$lWorkflow` — the workflow list used
- `$summary`   — lightweight metadata for the snapshot

```{r analytics}
snap_analytics <- study$analytics[["2012-01-31"]]
names(snap_analytics)

metric_names <- names(snap_analytics$results)
cat("Metrics computed:", paste(metric_names, collapse = ", "), "\n")

kri <- snap_analytics$results[[metric_names[1]]]
names(kri)           # Analysis_Summary, Analysis_Flagged, Analysis_Analyzed
kri$Analysis_Summary
kri$Analysis_Flagged

total <- sum(sapply(study$analytics, function(s) length(s$results)))
cat("Total metric results across all snapshots:", total, "\n")
```

## Reporting Pipeline Results

```{r reporting}
names(study$reporting)

snap_reporting <- study$reporting[["2012-01-31"]]
names(snap_reporting)
```

## Stepwise Pipeline Execution

Build the study in stages to inspect data before running heavy pipelines.

```{r stepwise}
raw_study <- create_longitudinal_study(
  study_id      = "STEP-001",
  participants  = 100,
  sites         = 10,
  snapshots     = 4,
  interval      = "1 month",
  domains       = c("AE", "LB", "VISIT", "QUERY"),
  run_analytics = FALSE,
  run_reporting = FALSE,
  verbose       = TRUE
)

raw_study <- run_longitudinal_analytics(raw_study, verbose = TRUE)
raw_study <- run_longitudinal_reporting(raw_study, verbose = TRUE)

names(raw_study)  # now includes $analytics and $reporting
```

## Low-Level Config API

For full control over dataset counts, temporal settings, and which
analytics/reporting packages and workflows to run.

```{r config-api}
custom_study <- create_study_config(
  study_id            = "CUSTOM-001",
  participant_count   = 300,
  site_count          = 20,
  analytics_package   = "gsm.kri",
  analytics_workflows = c("kri0001", "kri0002", "kri0003"),
  reporting_package   = "gsm.reporting"
) |>
  set_temporal_config(
    start_date     = "2023-06-01",
    snapshot_count = 12,
    snapshot_width = "months"
  ) |>
  add_dataset_config("Raw_AE",    enabled = TRUE) |>
  add_dataset_config("Raw_LB",    enabled = TRUE) |>
  add_dataset_config("Raw_VISIT", enabled = TRUE) |>
  add_dataset_config("Raw_QUERY", enabled = TRUE) |>
  add_dataset_config("Raw_DATACHG", enabled = TRUE,
    count_formula = function(config, snapshot_idx = 1) {
      round(config$study_params$participant_count * 0.4 * snapshot_idx /
              config$temporal_config$snapshot_count)
    }
  )

str(custom_study$study_params)
str(custom_study$temporal_config)
names(custom_study$dataset_configs)

raw_data <- generate_raw_data_from_config(custom_study, verbose = TRUE)
analytics <- generate_analytics_layers(raw_data, custom_study, verbose = TRUE)
reporting <- generate_reporting_layers(analytics, custom_study, verbose = TRUE)

study_obj <- create_longitudinal_study_data(
  study_id = custom_study$study_params$study_id,
  raw_data = raw_data,
  config = list(
    participants = custom_study$study_params$participant_count,
    sites        = custom_study$study_params$site_count,
    snapshots    = custom_study$temporal_config$snapshot_count,
    interval     = paste(1, custom_study$temporal_config$snapshot_width),
    domains      = gsub("^Raw_", "", names(custom_study$dataset_configs))
  )
)
study_obj$analytics <- analytics
study_obj$reporting <- reporting

summarize_longitudinal_study(study_obj)
```

## Exporting Study Data

`export_study_data()` writes a completed study to a structured folder
hierarchy on disk.

```
<output_dir>/<study_id>/
  <snapshot_date>/
    raw/          # Raw_*.csv
    mapped/       # Mapped_*.csv     (when analytics ran)
    analytics/    # <metric>_<table>.csv  (when analytics ran)
    reporting/    # Reporting_*.csv  (when reporting ran)
```

```{r export}
study_path <- export_study_data(
  study      = study,
  output_dir = tempdir(),
  verbose    = TRUE
)

# Parquet format (requires arrow package)
export_study_data(
  study        = study,
  output_dir   = ".",
  format       = "parquet",
  study_folder = "DEMO-001-export",
  overwrite    = TRUE
)

# Save full analytics object as RDS alongside CSVs
rds_path <- export_study_data(
  study      = study,
  output_dir = tempdir(),
  overwrite  = TRUE,
  save_rds   = TRUE
)

snap_date      <- names(study$raw_data)[1]
analytics_snap <- readRDS(file.path(rds_path, snap_date, "analytics_full.rds"))
names(analytics_snap)   # $results, $mapped, $lWorkflow, $summary
```

## Creating Multiple Studies

```{r multiple}
portfolio <- create_multiple_longitudinal_studies(
  study_names   = c("TRIAL-001", "TRIAL-002", "TRIAL-003"),
  participants  = 150,
  sites         = 12,
  snapshots     = 4,
  domains       = c("AE", "LB", "VISIT"),
  run_analytics = FALSE,
  verbose       = TRUE
)

names(portfolio)
length(portfolio[["TRIAL-001"]]$raw_data)   # 4 snapshots

portfolio_mixed <- create_multiple_longitudinal_studies(
  study_names  = c("PHASE2-SMALL", "PHASE3-LARGE"),
  participants = c(50, 300),
  sites        = c(5, 20),
  snapshots    = c(3, 12),
  study_configs = list(
    "PHASE2-SMALL" = list(domains = c("AE", "LB")),
    "PHASE3-LARGE" = list(domains = c("AE", "LB", "VISIT", "PD", "PK"))
  ),
  run_analytics = FALSE,
  verbose      = TRUE
)

print(portfolio_mixed)
summary(portfolio_mixed)

# Export all studies as parquet
export_study_data(
  study      = portfolio,
  output_dir = ".",
  format     = "parquet",
  overwrite  = TRUE,
  verbose    = TRUE
)
```
